---
title: "Introduction to ggdiceplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggdiceplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

```{r setup}
library(ggdiceplot)
library(ggplot2)
library(dplyr)
```

## Introduction

The `ggdiceplot` package provides ggplot2 extensions for creating dice-based visualizations. The main function, `geom_dice()`, displays data points as dice with 1-6 dots arranged in traditional dice patterns. This is particularly useful for visualizing categorical data with multiple dimensions in an intuitive and visually appealing way.

## Basic Usage

The most basic use of `geom_dice()` requires three aesthetic mappings:

- `x`: x-axis position
- `y`: y-axis position  
- `z`: number of dots on dice (1-6)

Let's create some sample data to demonstrate:

```{r basic-data}
# Create sample data
set.seed(123)
data <- expand.grid(x = 1:4, y = 1:3) %>%
  mutate(
    z = sample(1:6, n(), replace = TRUE),
    category = sample(c("Type A", "Type B", "Type C"), n(), replace = TRUE)
  )

head(data)
```

### Basic Dice Plot

```{r basic-plot}
# Basic dice plot
ggplot(data, aes(x = x, y = y, z = z)) +
  geom_dice() +
  scale_x_continuous(breaks = 1:4) +
  scale_y_continuous(breaks = 1:3) +
  labs(
    title = "Basic Dice Plot",
    subtitle = "Each dice shows 1-6 dots based on the z variable",
    x = "X Position",
    y = "Y Position"
  ) +
  theme_minimal()
```

## Adding Colors

You can color the dice dots using the `color` aesthetic:

```{r colored-plot}
# Dice plot with colors
ggplot(data, aes(x = x, y = y, z = z, color = category)) +
  geom_dice(dice_size = 1.2) +
  scale_color_manual(values = c("Type A" = "#E74C3C", "Type B" = "#3498DB", "Type C" = "#2ECC71")) +
  scale_x_continuous(breaks = 1:4) +
  scale_y_continuous(breaks = 1:3) +
  labs(
    title = "Colored Dice Plot",
    subtitle = "Dice dots colored by category",
    x = "X Position",
    y = "Y Position",
    color = "Category"
  ) +
  theme_minimal()
```

## Customizing Appearance

The `geom_dice()` function provides several parameters to customize the appearance:

```{r custom-styling}
# Custom styled dice plot
ggplot(data, aes(x = x, y = y, z = z, color = category)) +
  geom_dice(
    dice_size = 1.5,
    dice_color = "lightgray",
    dice_alpha = 0.9,
    dot_size = 0.4,
    dot_stroke = 0.8
  ) +
  scale_color_manual(values = c("#E74C3C", "#3498DB", "#2ECC71")) +
  scale_x_continuous(breaks = 1:4) +
  scale_y_continuous(breaks = 1:3) +
  labs(
    title = "Custom Styled Dice Plot",
    subtitle = "Custom dice background, dot size, and colors",
    x = "X Position",
    y = "Y Position",
    color = "Category"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#F8F9FA", color = NA)
  )
```

## Dice Patterns

The dice patterns follow traditional dice arrangements:

```{r dice-patterns}
# Show all dice patterns
pattern_data <- data.frame(
  x = rep(1:6, each = 1),
  y = rep(1, 6),
  z = 1:6,
  label = paste("Dice", 1:6)
)

ggplot(pattern_data, aes(x = x, y = y, z = z)) +
  geom_dice(dice_size = 2) +
  geom_text(aes(label = label), y = 0.5, size = 3) +
  scale_x_continuous(breaks = 1:6, limits = c(0.5, 6.5)) +
  scale_y_continuous(limits = c(0, 2)) +
  labs(
    title = "Traditional Dice Patterns",
    subtitle = "Standard arrangements for 1-6 dots",
    x = "Dice Value",
    y = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

## Advanced Usage

### Combining with Other Geoms

You can combine `geom_dice()` with other ggplot2 geoms:

```{r combined-geoms}
ggplot(data, aes(x = x, y = y)) +
  geom_tile(aes(fill = category), alpha = 0.3) +
  geom_dice(aes(z = z, color = category), dice_size = 0.8) +
  scale_fill_manual(values = c("#FFE5E5", "#E5F2FF", "#E5FFE5")) +
  scale_color_manual(values = c("#CC0000", "#0066CC", "#00CC00")) +
  scale_x_continuous(breaks = 1:4) +
  scale_y_continuous(breaks = 1:3) +
  labs(
    title = "Dice Plot with Background Tiles",
    x = "X Position",
    y = "Y Position"
  ) +
  theme_minimal()
```

### Faceting

Use faceting to create multiple dice plots:

```{r faceting}
ggplot(data, aes(x = x, y = y, z = z)) +
  geom_dice(dice_size = 0.8) +
  facet_wrap(~category) +
  scale_x_continuous(breaks = 1:4) +
  scale_y_continuous(breaks = 1:3) +
  labs(
    title = "Faceted Dice Plot",
    subtitle = "Separate plots for each category",
    x = "X Position",
    y = "Y Position"
  ) +
  theme_minimal()
```

## Real-World Example

Here's a more realistic example using dice plots to visualize survey responses:

```{r real-world}
# Create survey data
set.seed(456)
survey_data <- expand.grid(
  question = paste("Q", 1:5, sep = ""),
  respondent_group = c("Group A", "Group B", "Group C")
) %>%
  mutate(
    satisfaction = sample(1:6, n(), replace = TRUE, prob = c(0.05, 0.1, 0.15, 0.25, 0.25, 0.2)),
    x = as.numeric(factor(question)),
    y = as.numeric(factor(respondent_group))
  )

ggplot(survey_data, aes(x = x, y = y, z = satisfaction, color = respondent_group)) +
  geom_dice(dice_size = 1.2, dice_alpha = 0.7) +
  scale_color_manual(values = c("#FF6B6B", "#4ECDC4", "#45B7D1")) +
  scale_x_continuous(breaks = 1:5, labels = paste("Q", 1:5, sep = "")) +
  scale_y_continuous(breaks = 1:3, labels = c("Group A", "Group B", "Group C")) +
  labs(
    title = "Survey Satisfaction Scores",
    subtitle = "Dice dots represent satisfaction level (1-6)",
    x = "Survey Question",
    y = "Respondent Group",
    color = "Group"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Conclusion

The `ggdiceplot` package provides an intuitive way to visualize categorical data using dice representations. The `geom_dice()` function integrates seamlessly with the ggplot2 ecosystem, allowing for complex visualizations while maintaining the familiar ggplot2 syntax.

Key benefits:
- Intuitive visual representation using familiar dice patterns
- Full integration with ggplot2 aesthetics and themes
- Customizable appearance and styling
- Works with faceting and other ggplot2 features
- Suitable for various types of categorical data visualization